---
title: "Statistical Rethinking 2026, A04"
description: "My solution to exercise A04 from Richard McElreath's Statistical Rethinking course."
author: Paulina Sell
date: January 2026
date-format: MMM YYYY
categories: [R, Bayes, causal inference, prior predictive simulation, synthetic data]
toc: true
toc-depth: 4
image: false
execute:
  freeze: auto
---

[Link to full course on GitHub](https://github.com/rmcelreath/stat_rethinking_2026/tree/main) \
[Link to online lecture](https://youtu.be/GIdwLrW2nNo?si=VzBAB29ttkC8Cb7Q)

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
```

::: {.callout-tip  icon=false}
# Exercise
Using the Howell1 dataset, consider only the people younger than 13 years old. 
In this sample, estimate the causal effect of each month of growth on weight. 
Be sure to perform a prior predictive simulation to check and justify your priors. 
Also try to perform a validation of your model on synthetic data.
:::

## Prepare data

First, let's load the package and data.
Then, we'll keep only those younger than 13 and multiply age by 12 to get age in months.

```{r}
#| warning: false 
#| message: false 

library(rethinking)
data(Howell1)
d <- Howell1

d <- d[d$age < 13, ]
d$age_months <- d$age * 12
```

Now let's make a simple plot and visualize the weight of the children by age.

```{r}
#| code-fold: true

plot(
  weight ~ age_months,
  data = d,
  col = "red",
  xlab = "age in months",
  ylab = "weight in kg"
)
```

## Statistical model

We will rely on a linear regression that predicts average weight (kg) depending on age (months). 

This is the model:

$$
\text{weight} \sim \mathcal{N}(\mu, \sigma)
$$
$$
\mu = \alpha + \beta (\text{age}_i - \overline{age})
$$ 

Now we can set some priors.
We'll start with very standard / boring priors.

$$
\alpha \sim \mathcal{N}(0, 10)
$$
$$
\beta \sim \text{Uniform}(0, 1)
$$

$$
\sigma \sim \text{Uniform}(0, 5)
$$

## Prior predictive simulation
Let's plot some lines from our model before letting it look at the data: 

```{r}
n <- 1000
a <- rnorm(n, 0, 10)
b <- runif(n, 0, 1)

plot(
  NULL,
  xlim = c(0, 180),
  ylim = c(0, 40),
  xlab = "age (months)",
  ylab = "weight (kg)"
)
mtext("a ~ dnorm(0,10), b ~ dunif(0,1)")

for (j in 1:100) {
  abline(a = a[j], b = b[j], lwd = 2, col = 2)
}
```

This is not perfect, but as McElreath said in his lecture: "There are no correct priors, only scientifically justifiable priors".

## Simulation-based validation
Let's now generate synthetic data from our model and see if it makes any sense.

### Simulate data
We need a function to generate age and weight data of hypothetical people.

```{r}
sim_age_weight <- function(N, b, a) {
  age_months <- runif(N, 0, 13) * 12
  weight <- a + b * age_months + rnorm(N, 0, 2)
  data.frame(age_months, weight)
}

dat <- sim_age_weight(1000, 0.3, 0)

head(dat)
```

Let's have a look: 
```{r}
#| code-fold: true

plot(
  weight ~ age_months,
  data = dat,
  col = "red",
  xlab = "age (months)",
  ylab = "weight (kg)"
)
```

Okay, a bit too linear maybe, but that's fine.
Nothing impossible!


### Fit model
Let's fit our model to this synthetic data:

```{r}
age_mean <- mean(dat$age_months)

model_synth <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + b * (age_months - age_mean),
    a ~ dnorm(0, 10),
    b ~ dunif(0, 1),
    sigma ~ dunif(0, 5)
  ),
  data = dat
)

p <- precis(model_synth)
p
```

Okay so we've successfully recovered `b = 0.3` from our data simulation, but what about `a`?
We've set `a = 0` in the simulation, but now it's `r round(p$mean[1], 2)`? \
Fortunately that's not a bug! 
`sim_age_weight()` defines `a` as weight at `age = 0`, but our fitted model centers on `age_mean`, so the fitted `a` is weight at the mean age instead.
`b` doesn't care about centering, so it comes back correctly. \
We can do a quick check: 

```{r}
0.3 * mean(dat$age_months)
```

Let's plot the simulated data (red) & the posterior mean regression line (black): 

```{r}
#| code-fold: true

post_rand <- extract.samples(model_synth, n = 20)
post_full <- extract.samples(model_synth)
a_map <- mean(post_full$a)
b_map <- mean(post_full$b)

plot(
  weight ~ age_months,
  data = dat,
  col = "red",
  xlab = "age (months)",
  ylab = "weight (kg)"
)
mtext("N = 157")

for (i in 1:20) {
  curve(
    post_rand$a[i] + post_rand$b[i] * (x - age_mean),
    col = col.alpha("grey", 0.4),
    add = TRUE
  )
}

curve(a_map + b_map * (x - age_mean), add = TRUE, col = "black", lwd = 2)
```

Great!! :)

## Fit the model to the real data

For the regression model, we also need the mean age:

```{r}
age_mean <- mean(d$age_months)
```

### Fit model
Let's try and fit that model using the `rethinking` package! 

```{r}
#| label: mA4_quap

mA4 <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + b * (age_months - age_mean),
    a ~ dnorm(0, 10),
    b ~ dunif(0, 1),
    sigma ~ dunif(0, 5)
  ),
  data = d
)

precis(mA4)
```

### Visualize
Okay, now let's look at the data and the line at the posterior mean plotted (black) and 20 samples from the posterior (grey):
```{r}
#| code-fold: true

post_rand <- extract.samples(mA4, n = 20)
post_full <- extract.samples(mA4)
a_map <- mean(post_full$a)
b_map <- mean(post_full$b)

plot(
  weight ~ age_months,
  data = d,
  col = "red",
  xlab = "age in months",
  ylab = "weight in kg"
)
mtext("N = 157")

for (i in 1:20) {
  curve(
    post_rand$a[i] + post_rand$b[i] * (x - age_mean),
    col = col.alpha("grey", 0.4),
    add = TRUE
  )
}

curve(a_map + b_map * (x - age_mean), add = TRUE, col = "black", lwd = 2)
```

If we had less data (n = 20), then the uncertainty would be larger: 

```{r}
#| code-fold: true

d_small <- d[1:20, ]
mean_age_small <- mean(d_small$age_months)

mA4_small <- quap(
  alist(
    weight ~ dnorm(mu, sigma),
    mu <- a + b * (age_months - mean_age_small),
    a ~ dnorm(0, 10),
    b ~ dunif(0, 1),
    sigma ~ dunif(0, 5)
  ),
  data = d_small
)

post_rand <- extract.samples(mA4_small, n = 20)
post_full <- extract.samples(mA4_small)
a_map <- mean(post_full$a)
b_map <- mean(post_full$b)

plot(
  weight ~ age_months,
  data = d_small,
  col = "red",
  xlab = "age in months",
  ylab = "weight in kg"
)

for (i in 1:20) {
  curve(
    post_rand$a[i] + post_rand$b[i] * (x - mean_age_small),
    col = col.alpha("grey", 0.4),
    add = TRUE
  )
}

curve(a_map + b_map * (x - mean_age_small), col = "black", lwd = 2, add = T)

mtext("N = 20")
```

So, what's the answer?
For children under 13, each extra month of age adds about [b] kg of weight on average (89% CI: [lower, upper]). The synthetic data check gave us confidence the model recovers the true slope, and the n = 20 version shows what happens with less data: same trend, way more uncertainty. A straight line is obviously a simplification — growth isn't linear from infancy to age 13! — but it does the job here.