---
title: "Comparing Bayesian and frequentist meta-analysis in R"
description: ""
date: July 2026
date-format: "MMM YYYY"
categories: [R, meta-analysis, brms, metafor, simulation]
image: false
toc: true
toc-depth: 4
execute:
  cache: true
  freeze: auto
---

# Simulating studies

First, we need to simulate some studies that we can then pool.

## Study sizes

All of them need a sample size, effect estimate, standard error (SE) and ideally a first author :`).

We begin with the sample size.

```{r}
#| label: load_packages
#| message: false
#| warning: false

library(brms)
library(metafor)
library(tidyverse)
library(tidybayes)
library(ggridges)
library(glue)
library(here)
library(magick)
```

```{r}
#| label: sim_sizes

n <- 15
weights <- c(0.4, 0.5, 0.1)

components <- sample(1:3, size = n, prob = weights, replace = TRUE)

possible_sizes <- c(50, 500, 1000)
sample_size <- rpois(n, lambda = possible_sizes[components])
```

We first decide that this meta-analysis will consist of `r n` studies.
We want to sample sample sizes out of a mixed distribution (that just means we mix multiple distributions), because just sampling out of something like a single normal distribution would be kind of boring.. all studies would have similar sample sizes, which is something that we rarely see in reality. 
`components` becomes a vector with length `r n` that gets randomly filled with `1`, `2` and `3`, depending on the probability that we set in `weights` (e.g. `1` has a probability of `r weights[1] * 100`%).
Then, we decide for approximate sample sizes and save them in a simple vector `possible_sizes`.
I've chosen `r possible_sizes[1]`, `r possible_sizes[2]` and `r possible_sizes[3]`, so that most studies are small or medium-sized and some (`r weights[3] * 100`%) are really large.

To introduce some randomness, we sample from a poisson distribution (great for counts) `r n` times and each time we get the correct approximate sample size (from `possible_sizes`) for the `components` vector by indexing.

I don't know about you, but I found this to be pretty confusing and not very intuitive.
I promise the next steps will we way more straightforward.

## Effect sizes

Now we sample effect sizes from a normal distribution.

```{r}
#| label: sim_effect-sizes

effect_size <- rnorm(n, 1.9, 1)
```

## Standard error

To simulate standard errors (SE) for our studies, we just sample standard deviations from an uniform distribution and then simply use the formula to calculate SE.

```{r}
#| label: sim_SE

# SD <- rexp(n, 0.8)
SD <- rcauchy(n, location = 3, scale = 0.2)

SE <- SD / sqrt(sample_size)
```

We're almost ready! 
Here I'm making up some author names and then stich together a nice tibble with all the data we need.
```{r}
#| label: build_studies_df
#| code-fold: true

library(tibble)

authors <- c(
  "Gustav et al",
  "Amanda and Osti",
  "Vengur et al",
  "Bjalla et al",
  "Flugsvinn et al",
  "Fiedel et al",
  "Mimi and Bobby",
  "Kapa et al",
  "Lilly et al",
  "Trausti et al",
  "Flocke and Buddy",
  "Beethoven et al",
  "Olli et al",
  "Bekecs et al",
  "Dalia et al"
)

dat <- tibble(
  authors,
  effect_size,
  SE,
  sample_size
)

dat[1:15, ]
```


# Statistical model

## The model

We model each study's observed effect $\hat{\theta}_{k}$ as a noisy reflection of the true effect in that study’s sample $\theta_{k}$, with the noise represented by the within-study sampling error $\sigma_{k}$ ​(treated as known based on reported standard errors). 
These true effects $\theta_{k}$ are themselves modelled as draws from an overarching normal distribution which describes the true average effect across studies $\mu$ and the between-study heterogeneity $\tau$.

$$
\begin{split}
\hat{\theta}_{k} &\sim \mathcal{N}(\theta_{k}, \sigma_{k}) \\
\theta_{k} &\sim \mathcal{N}(\mu, \tau) \\
\end{split}
$$

Where $\hat{\theta}_{k}$ is the observed effect size for study $k$, $\sigma_{k}$ is the within-study sampling standard deviation for study $k$ (treated as known from the reported SE of each study), $\theta_{k}$ is the true underlying effect for study $k$, $\mu$ is the overall pooled effect and $\tau$ is the between-study standard deviation (heterogeneity).

# Bayesian meta-analysis

## Priors

```{r}
#| label: def_priors
#| echo: false

prior_mu_mean <- 2
prior_mu_sd <- 1

prior_tau_mean <- 0
prior_tau_sd <- 2
```

$\mu$ and $\tau$ are the parameters we'll actually get posterior distributions for.
$\mu$ describes the pooled effect across all studies and $\tau$ describes the between-study heterogeneity.
Now we need to set some priors for these two parameters.

<!-- add info about priors for beginners? -->

### Mu
What effect size could we reasonably expect here?
We could of course be lazy and just use parameters we set in the simulation above, but that's lame (and in reality we of course don't have that knowledge).\
Let's say that from the literature we've read so far and plausibility, we are comfortable to say that the effect is probably positive not super big.
We can translate this into $\mu$ ~ $\mathcal{N}(`r prior_mu_mean`, `r prior_mu_sd`)$.

```{r}
#| label: vis_prior_mu
#| code-fold: true

x <- seq((-1), (5), by = 0.1)
y <- dnorm(x, prior_mu_mean, prior_mu_sd)
plot(x, y, type = "l", xlab = "true effect", ylab = "", yaxt = "n")

lower_bound <- 0
upper_bound <- 4

idx <- x >= lower_bound & x <= upper_bound
x_shade <- x[idx]
y_shade <- y[idx]

polygon(
  c(x_shade[1], x_shade, tail(x_shade, 1)),
  c(0, y_shade, 0),
  col = adjustcolor("cornflowerblue", alpha.f = 0.5),
  border = NA
)
```

With this prior, we say that we think the true effect is somewhere between `r lower_bound` and `r upper_bound` with a probability of `r round((pnorm(4, 2, 1) - pnorm(0, 2, 1)) * 100)`%.

### Tau

We also have to set a prior for $\tau$, tau, the between-study heterogeneity.
Heterogeneity can never be negative and we don't want to be super restrictive with this prior.
We could use something like a cauchy distribution here, but I'll just do a half-normal, meaning that we clip the normal distribution at zero.
We'll set the prior for tau to $\tau \sim \mathcal{N}^+(`r prior_tau_mean`, `r prior_tau_sd`)$

```{r}
#| label: vis_prior_tau
#| code-fold: true

x <- seq((0), (6), by = 0.1)
y <- dnorm(x, 0, 2)
plot(
  x,
  y,
  type = "l",
  xlab = "between-study heterogeneity",
  ylab = "",
  yaxt = "n"
)

lower_bound <- 0
upper_bound <- 3

idx <- x >= lower_bound & x <= upper_bound
x_shade <- x[idx]
y_shade <- y[idx]

polygon(
  c(x_shade[1], x_shade, tail(x_shade, 1)),
  c(0, y_shade, 0),
  col = adjustcolor("orange", alpha.f = 0.3),
  border = NA
)
```

Here we say: "The probability, that the true between-study heterogeneity is smaller than `r upper_bound` is `r round((1 - (1 - pnorm(upper_bound, prior_tau_mean, prior_tau_sd)) * 2) * 100)`%".

Now we save our priors:

```{r}
#| label: save_priors
#| eval: false

prior_mu_mean <- 2
prior_mu_sd <- 1

prior_tau_mean <- 0
prior_tau_sd <- 2
```

This is the full statistical model, including both priors:
$$
\begin{split}
\hat{\theta}_{k} &\sim \mathcal{N}(\theta_{k}, \sigma_{k}) \\
\theta_{k} &\sim \mathcal{N}(\mu, \tau) \\
\mu &\sim \mathcal{N}(`r prior_mu_mean`, `r prior_mu_sd`) \\
\tau &\sim \mathcal{N^+}(`r prior_tau_mean`, `r prior_tau_sd`)
\end{split}
$$ {#eq-model}


## Fitting the model

Okay, now we use `brms`'s `prior` function to define the priors for the `brms` code.

```{r}
#| label: priors

priors <- c(
  prior(normal(prior_mu_mean, prior_mu_sd), class = Intercept),
  prior(normal(prior_tau_mean, prior_tau_sd), class = sd, lb = 0)
)
```

Behind the doors `brms` code gets compiled into Stan code.
Because Stan needs a little help to understand that `prior_mu_mean` means `r prior_mu_mean`, we save the names into a `stanvar` object. 
You don't need to worry about this if you just instert numbers directly into the code above.

```{r}
#| label: stanvars
#| code-fold: true

sv <-
  stanvar(prior_mu_mean, name = "prior_mu_mean") +
  stanvar(prior_mu_sd, name = "prior_mu_sd") +
  stanvar(prior_tau_mean, name = "prior_tau_mean") +
  stanvar(prior_tau_sd, name = "prior_tau_sd")
```

Okay, now we are ready to write down the `brms` model.

```{r}
#| label: main_model
#| output: false

# Main model
m.brm <- brm(
  effect_size | se(SE) ~ 1 + (1 | authors),
  data = dat,
  prior = priors,
  stanvars = sv,
  save_pars = save_pars(all = TRUE),
  chains = 4,
  iter = 4000
)
```

The second line is our model formula. 
`effect_size | se(SE)` captures $\hat{\theta}_k \sim \mathcal{N}(\theta_k, \sigma_k)$, each study's observed effect together with its known standard error;
`~ 1` is the pooled effect $\mu$;
and `+ (1 | authors)` gives each study its own true effect $\theta_k$, rather than forcing them all to share exactly one value.\
With `save_pars` we make sure that all parameters get saved, which `brms` doesn't do by default.
We run 4 MCMC chains, so that we can check convergence — if all 4 chains mix well, that's a good sign the sampler explored the posterior properly rather than getting stuck somewhere.

Next, we fit the model again and only sample from the prior so that we can check if our priors are nonsense or not.

```{r}
#| label: sample_from_prior
#| output: false

# Sample from prior only
fitPrior <- brm(
  effect_size | se(SE) ~ 1 + (1 | authors),
  data = dat,
  prior = priors,
  stanvars = sv,
  sample_prior = "only",
  chains = 4,
  iter = 4000
)
```

## Diagnostics

Okay, now we do some prior and posterior predictive checks and plot the posterior distributions and MCMC chains.
```{r}
#| label: spread_draws

draws <- spread_draws(m.brm, b_Intercept, sd_authors__Intercept)
pp_check(fitPrior, ndraws = 20)
pp_check(m.brm, ndraws = 20)
plot(m.brm, variable = c("b_Intercept", "sd_authors__Intercept"))
```

Looks good! 

Now let's have a look at the result summary:

```{r}
#| label: brms_results

summary(m.brm)
```

# Frequentist meta-analysis

We also fit a frequentist model using the `metafor` package.

```{r}
#| label: freq_model

m.freq <- rma(
  yi = effect_size,
  sei = SE,
  data = dat,
  method = "REML"
)

m.freq
```

# Forest plots

```{r}
#| label: data_wrangling

study.draws <- spread_draws(
  m.brm,
  r_authors[authors, ],
  b_Intercept
) %>%
  mutate(b_Intercept = r_authors + b_Intercept)

pooled.effect.draws <- spread_draws(m.brm, b_Intercept) %>%
  mutate(authors = "Pooled Estimate")

forest.data <- bind_rows(study.draws, pooled.effect.draws) %>%
  ungroup() %>%
  mutate(authors = str_replace_all(authors, "[.]", " ")) %>%
  mutate(
    authors = factor(
      authors,
      levels = c(rev(dat$authors), "Pooled Estimate")
    )
  )

forest.data.summary <- group_by(forest.data, authors) %>%
  mean_qi(b_Intercept) # calc mean & 2.5 & 97.5 Q
```

```{r}
#| label: ordering_studies
#| echo: false

n_studies <- nlevels(forest.data$authors) # includes pooled estimate

study_levels <- setdiff(levels(forest.data$authors), "Pooled Estimate")

new_levels <- c(
  "Pooled Estimate",
  "spacer_bottom",
  study_levels,
  "spacer_top"
)

forest.data <- forest.data %>%
  mutate(authors = factor(authors, levels = new_levels))

forest.data.summary <- forest.data.summary %>%
  mutate(authors = factor(as.character(authors), levels = new_levels))

# Append NA rows for the two spacer positions
spacer_rows <- tibble(
  authors = factor(c("spacer_bottom", "spacer_top"), levels = new_levels),
  b_Intercept = NA_real_,
  .lower = NA_real_,
  .upper = NA_real_
)
forest.data.summary <- bind_rows(forest.data.summary, spacer_rows)

n_levels_total <- n_studies + 2
```

```{r}
#| label: set_sizes
#| echo: false

linewidth <- 0.35
base_size_theme <- 12
base_size_geom <- base_size_theme * 0.352
title_size <- 14
```

```{r}
#| label: Bayes_forestplot
#| echo: false

# Bayes_forest_plot <-

Bayes_forest_plot <- ggplot(
  aes(b_Intercept, authors),
  data = forest.data
) +
  annotate(
    "segment",
    x = 0,
    xend = 0,
    y = 0,
    yend = n_levels_total,
    linewidth = linewidth,
    linetype = "dashed"
  ) +
  geom_density_ridges(
    fill = "grey",
    rel_min_height = 0.01,
    col = NA,
    scale = 1
  ) +
  geom_pointinterval(
    data = forest.data.summary %>% filter(!is.na(b_Intercept)),
    aes(xmin = .lower, xmax = .upper),
    shape = 22,
    point_fill = "black",
    fatten_point = 1.5,
    linewidth = 0.15
  ) +
  geom_text(
    data = mutate_if(
      forest.data.summary %>% filter(!is.na(b_Intercept)),
      is.numeric,
      round,
      2
    ),
    aes(
      label = glue(
        "{format(b_Intercept, nsmall = 2)} [{format(.lower, nsmall = 2)}, {format(.upper, nsmall = 2)}]"
      ),
      x = 4.2
    ),
    hjust = 0,
    size = base_size_geom
  ) +
  annotate(
    # "Study" header
    "text",
    x = -1.4,
    y = n_levels_total + 1,
    label = "Study",
    hjust = 1.05,
    fontface = "bold",
    size = base_size_geom
  ) +
  annotate(
    # "Estimate [95% CrI]" header
    "text",
    x = 4.5,
    y = n_levels_total + 1,
    label = "Estimate [95% CrI]",
    hjust = 0.27,
    fontface = "bold",
    size = base_size_geom
  ) +
  annotate(
    # horizontal line top
    "segment",
    x = -5,
    xend = 8,
    y = n_levels_total,
    yend = n_levels_total,
    linewidth = linewidth
  ) +
  annotate(
    # horizontal line bottom
    "segment",
    x = -5,
    xend = 8,
    y = 2,
    yend = 2,
    linewidth = linewidth
  ) +
  labs(
    title = "Bayesian meta-analysis (ggplot2)",
    x = "Effect",
    y = NULL
  ) +
  scale_y_discrete(
    expand = expansion(add = c(0.5, 2.4)),
    labels = function(x) ifelse(x %in% c("spacer_bottom", "spacer_top"), "", x),
    drop = FALSE
  ) +
  scale_x_continuous(
    breaks = seq(-5, 4, by = 4),
    expand = expansion(add = c(1, 1))
  ) +
  coord_cartesian(
    xlim = c(0, 5),
    clip = "off"
  ) +
  guides(x = guide_axis(cap = "both")) +
  theme_minimal() +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.margin = margin(t = 18.5, r = 10, b = 35, l = 20),
    plot.title = element_text(hjust = 0, face = "bold", size = title_size),
    plot.background = element_rect(fill = "#FBFAF9", colour = NA),
    axis.text.y = element_text(
      hjust = 0,
      margin = margin(r = -10),
      size = base_size_theme,
      colour = "black"
    ),
    axis.text.x = element_text(vjust = -1, size = base_size_theme),
    axis.title.x = element_text(vjust = -2.7, size = base_size_theme),
    axis.line.x = element_line(linewidth = linewidth),
    axis.ticks.x = element_line(linewidth = linewidth),
    axis.ticks.length.x = unit(0.3, "cm"),
  )
```

```{r}
#| label: compare
#| code-fold: true

ggsave(
  "data/Bayes_forest_plot.png",
  Bayes_forest_plot,
  width = 20,
  height = 15,
  units = "cm",
  bg = "#FBFAF9"
)

png(
  "data/freq_forestplot.png",
  width = 20,
  height = 15,
  units = "cm",
  res = 300,
  bg = "#FBFAF9"
)

forest(
  freq_model,
  slab = str_replace_all(freq_model$data$author_year, "[.]", " "),
  mlab = "Pooled Estimate",
  xlab = "Effect",
  main = "Frequentist meta-analysis (metafor)"
)

dev.off()

img1 <- image_read("data/Bayes_forest_plot.png")
img2 <- image_read("data/freq_forestplot.png")
combined <- image_append(c(img1, img2), stack = FALSE)
image_write(combined, "data/forest_plots_combined.png")
```