---
title: "Statistical Rethinking 2026, A08"
description: "My solution to exercise A08 from Richard McElreath's Statistical Rethinking course."
author: Paulina Sell
date: March 2026
date-format: MMM YYYY
categories: [R, Bayes, causal inference, MCMC]
toc: true
toc-depth: 4
image: false
execute:
  freeze: auto
---

[Link to full course on GitHub](https://github.com/rmcelreath/stat_rethinking_2026/tree/main) \
[Link to online lecture](https://youtu.be/N_LRQUrdHag?si=dN_uS8JvjWGmWHnY)

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))
```

::: {.callout-tip  icon=false}
## Exercise
Modify the item response model from the lecture to ask if there are any average differences between French and American judges. 
Use `ulam()` and be sure to check chain convergence. 
Do the French and American judges differ in either their harshness or their discrimination?
:::

First, we load the package and data.

```{r}
#| label: load_pck_dat
#| message: false
#| warning: false

library(rethinking)
data(Wines2012)
data <- Wines2012
```

Let's have a look: 

```{r}
#| label: view_data
#| echo: false

str(data)
```

The data is from a wine-tasting event, where 20 wines (10 from France, 10 from New Jersey) have been tasted by 9 French and American judges.\
Now, we want to find out whether there are any differences in harshness or discrimination between French and American judges.
Let's use a DAG to think about this.

```{r}
#| label: DAG1
#| echo: false
#| fig-height: 1.5
#| fig-width: 3
#| fig-align: center
#| message: false
#| warning: false

library(dagitty)
library(ggdag)
library(ggplot2)

dag <- dagitty(
  "dag {
   Q -> S
   X -> Q
   X -> S
   J -> S
   Z -> J
}"
)

coordinates(dag) <- list(
  x = c(Q = 1, X = 1, S = 2, J = 2, Z = 3),
  y = c(Q = 2, X = 1, S = 2, J = 1, Z = 1)
)

ggdag_classic(dag) +
  theme_dag() + 
  theme(
    plot.background = element_rect(fill = "transparent", color = NA))
```

Q = Wine quality\
S = Score\
X = Wine origin\
J = Judge preferences\
Z = Judge origin

## Statistical Model

The score $S$ assigned to a wine is modeled as normally distributed around a mean $\mu$:

$$
\text{S} \sim \mathcal{N}(\mu, \sigma)
$$
$$
\mu = (\text{Q}_W + \text{O}_X - \text{H}_Z) * \text{D}_Z
$$

Inside the brackets, three additive components determine the underlying score before judge effects are applied: $\text{Q}_W$ is the intrinsic quality of wine $\text{W}$; $\text{O}_X$​ is a bonus or penalty depending on whether the wine's origin $\text{X}$ matches the judge's prior expectations; and $\text{H}_Z$ is the harshness of judge $\text{Z}$, which shifts scores downward.
This sum is then multiplied by $\text{D}_Z$​, the discrimination of judge $\text{Z}$. 

## Priors 
I don't know what to expect from the model yet, so I'll first set some default priors but I know that discrimination ($\text{D}$) and $\sigma$ must be positive.

$$
\text{Q} \sim \mathcal{N}(0, 1)
$$
$$
\text{O} \sim \mathcal{N}(0,1)
$$
$$
\text{H} \sim \mathcal{N}(0,1)
$$
$$
\text{D} \sim \text{exponential}(1)
$$
$$
\sigma \sim \text{exponential}(1)
$$

## Prepare data
Now we need to prepare the data.
This includes standardizing the outcome variable, $\text{S}$.

```{r}
#| label: prepare_data

data <- list(
  S = standardize(data$score),
  W = data$wine,
  X = ifelse(data$wine.amer == 1, 1, 2), # 1 = american wine, 2 = french wine
  Z = ifelse(data$judge.amer == 1, 1, 2) # 1 = american judge, 2 = french judge
)
```

## Fit model
Let's try and fit the model:

```{r}
#| label: fit_ulam
#| warning: false
#| output: false

m.wines <- ulam(
  alist(
    S ~ dnorm(mu, sigma),
    mu <- (Q[W] + O[X] - H[Z]) * D[Z],
    Q[W] ~ dnorm(0, 1),
    O[X] ~ dnorm(0, 1),
    H[Z] ~ dnorm(0, 1),
    D[Z] ~ dexp(1),
    sigma ~ dexp(1)
  ),
  data = data, chains = 4
)
```

```{r}
#| label: diagnostics

show(m.wines)
precis(m.wines, depth = 2)
trankplot(m.wines)
```

## Causal contrast
### Total causal contrast in mean: wine discrimination

```{r}
#| label: causal_contrast_mean_D
#| code-fold: true

post <- extract.samples(m.wines)

mu_contrast <- post$D[, 2] - post$D[, 1] # french - american discrimination

dens(
  mu_contrast,
  lwd = 2,
  xlab = "posterior mean discrimination contrast"
)
```

### Total causal contrast in predicted score: wine discrimination

```{r}
#| label: causal_contrast_pred_D
#| code-fold: true

# sample from posterior
D_A <- rnorm(1000, post$D[, 1], post$sigma)
D_F <- rnorm(1000, post$D[, 2], post$sigma)

# calculate contrast
S_contrast <- D_F - D_A

dens(
  S_contrast,
  lwd = 2,
  xlab = "posterior contrast in discrimination (French - American)"
)
```

Here we can see that `r (sum(S_contrast > 0) / 1000) * 100`% of the times we randomly sample a new judge, the French judge is more discriminating than the American judge. 
And `r (sum(S_contrast < 0) / 1000) * 100`% of the time, the American judge is more discriminating than the French judge.
So it seems like there is no difference between American and French wine judges in terms of wine discrimination.\
Let's now look at harshness.

### Total causal contrast in mean: harshness

```{r}
#| label: causal_contrast_mean_H
#| code-fold: true

mu_contrast_H <- post$H[, 2] - post$H[, 1] # french - american discrimination

dens(
  mu_contrast_H,
  lwd = 2,
  xlab = "posterior mean harshness contrast"
)
```

### Total causal contrast in predicted score: harshness 

```{r}
#| label: causal_contrast_pred_H
#| code-fold: true

# sample from posterior
H_A <- rnorm(1000, post$H[, 1], post$sigma)
H_F <- rnorm(1000, post$H[, 2], post$sigma)

# calculate contrast
S_contrast_H <- H_F - H_A

dens(
  S_contrast_H,
  lwd = 2,
  xlab = "posterior contrast in harshness (French - American)"
)
```

Here we can see that `r (sum(S_contrast_H > 0) / 1000) * 100`% of the times we randomly sample a new judge, the French judge is more harsh in their judgement than the American judge. 
And `r (sum(S_contrast_H < 0) / 1000) * 100`% of the time, the American judge is more harsh in their judgement than the French judge.

## Lineplot
Now I want to try to recreate the lineplot that McElreath showed in his [slides](https://youtu.be/N_LRQUrdHag?si=xfDaVd0v7V-usOPV&t=3851).

First, we need to create a sequence of wine quality (Q).
```{r}
#| label: prepare_data_2
#| message: false

library(tidyverse)

# Simulation settings
n_lines <- 40
Q_seq <- seq(-2, 2, len = 60)
# 1 = American, 2 = French
Z_american <- 1
Z_french <- 2
```

Now, we need a function that gets the right values out of the posteriors. 
We first sample indices from the posterior, so that we "decide" a priori what row from the posterior we'll use.
In the `for` loop, we then fill the empty list we initiated with values from the full posterior in order to get the trajectories for a given Z.
```{r}
#| label: function

make_lines <- function(Z_val, n_lines, Q_seq, post) {
  idx <- sample(nrow(post$sigma), n_lines, replace = FALSE)

  results <- vector("list", length(idx))

  for (i in seq_along(idx)) {
    s <- idx[i] # Full posterior draw: O, H & D come from the same sample row
    X_draw <- sample(1:2, 1) # random wine origin
    O <- post$O[s, X_draw]
    H <- post$H[s, Z_val]
    D <- post$D[s, Z_val]

    results[[i]] <- data.frame(
      Q = Q_seq,
      mu = (Q_seq + O - H) * D,
      line = i
    )
  }

  # Combine all the individual data frames into one
  do.call(rbind, results)
}
```

Now, we just call the function twice (for French and American judges) and then combine both data frames the data by row-binding it.

```{r}
#| label: gather_data_in_df

df_plot <- bind_rows(
  make_lines(Z_american, n_lines, Q_seq, post) |>
    mutate(judge_origin = "American"),
  make_lines(Z_french, n_lines, Q_seq, post) |> 
    mutate(judge_origin = "French")
) |>
  mutate(judge_origin = factor(judge_origin, levels = c("American", "French")))
```

 And now we're ready to plot!
```{r}
#| label: plot

ggplot(df_plot, aes(x = Q, y = mu, group = line)) +
  geom_vline(xintercept = 0, col = "lightcoral") +
  geom_hline(yintercept = 0, col = "lightcoral") +
  geom_line(alpha = 0.35, linewidth = 0.5, colour = "#2c7bb6") +
  facet_wrap(~judge_origin, ncol = 2) +
  labs(
    x = "Wine quality (Q)",
    y = "Predicted score (μ)",
    title = "Predicted scores by judge origin",
    subtitle = paste0(n_lines, " posterior draws per panel")
  ) +
  theme_classic()
```

So: do the French and American judges actually differ in either their harshness or discrimination?
Yes, they do differ in harshness.
French judges tend to give lower scores in general compared to American judges.
But their discrimination between wines does not differ.