---
title: "Statistical Rethinking 2026, A06"
description: "Simulating synthetic data and using a DAG to think about causal inference."
author: Paulina Sell
date: February 2026
date-format: MMM YYYY
categories: [R, Bayes, DAG, causal inference, confounding, simulation]
toc: true
toc-depth: 4
image: false
execute:
  freeze: auto
---

::: {.callout-tip  icon=false}
# TL;DR
In this sixth exercise from the beginner section, we first simulate data based on a DAG (Directed Acyclic Graph) and then estimate the causal effect between two variables.
The cool thing is that at the end we can actually check if we were able to "recover" the relationship between the variables we set when simulating the data.
On the way to the right answer, we'll need to think about confounding and use some do-calculus.
:::

[Link to full course on GitHub](https://github.com/rmcelreath/stat_rethinking_2026/tree/main) \
[Link to online lecture](https://www.youtube.com/watch?v=lGR7D45Ww38)

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(dev.args = list(bg = "#FBFAF9"))
```

## Exercise
*"Consider the DAG below:*
```{r}
#| label: DAG1  
#| echo: false
#| fig-height: 1.5
#| fig-width: 3
#| fig-align: center
#| message: false
#| warning: false

library(dagitty)
library(ggdag)
library(HDInterval)
library(ggplot2)

dag <- dagitty(
  "dag {
    X -> Y
    Z -> X
    Z -> Y
    A -> Y
    A -> Z
}"
)

coordinates(dag) <- list(
  x = c(X = 1, Y = 3, Z = 2, A = 4),
  y = c(X = 1, Y = 1, Z = 0, A = 0)
)

ggdag_classic(dag) +
  theme_dag() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )
```

*I want you to first simulate data consistent with this DAG. 
You can assume that each variable is normally distributed.
Then focus on estimating the causal effect of X on Y. 
Which other variables are needed to correctly estimate the effect of X on Y? 
For any variables other than X, what is the meaning of their estimated coefficients? Are they also causal effect estimates?
Why or why not?"*

## Simulate data from DAG
First, let's simulate some data according to the DAG.
I found page 138 in the book helpful for this.

```{r}
#| label: simulate-data
#| cache: false

N <- 1000
b_XY <- 0.5
b_ZX <- 1
b_ZY <- -1
b_AZ <- 1
b_AY <- -1

set.seed(20)
A <- rnorm(N, 0, 1) # sim A
Z <- rnorm(N, b_AZ * A, 1) # sim A -> Z
X <- rnorm(N, b_ZX * Z, 1) # sim Z -> X
Y <- rnorm(N, b_XY * X + b_ZY * Z + b_AY * A, 1) # sim X -> Y; Z -> Y; A -> Y

df <- data.frame(A, Z, X, Y)
df[1:5, ]
```

This is a simple plot of the four variables we just generated: 

```{r}
#| label: plot_sim_data
#| message: false
#| code-fold: true

library(rethinking)

dens(df$A, col = "#E77D72", lwd = 2, lty = 2, xlim = c(-8, 5))
dens(df$Z, col = "#7291e7ff", lwd = 2, lty = 3, add = T)
dens(df$X, col = "#72e7b0ff", lwd = 2, lty = 4, add = T)
dens(df$Y, col = "black", lwd = 2, lty = 1, add = T)
legend(
  x = "topright",
  legend = c("A", "Z", "X", "Y"),
  lty = c(2, 3, 4, 1),
  lwd = 2,
  col = c("#E77D72", "#7291e7ff", "#72e7b0ff", "black")
)
```
## Total effect of X on Y

We have two stacked *forks* here:

```{r}
#| label: DAG2
#| echo: false
#| fig-height: 1.5
#| fig-width: 3
#| fig-align: center

dag <- dagitty(
  "dag {
    X -> Y
    Z -> X
    Z -> Y
    A -> Y
    A -> Z
}"
)

coordinates(dag) <- list(
  x = c(X = 1, Y = 3, Z = 2, A = 4),
  y = c(X = 1, Y = 1, Z = 0, A = 0)
)

ggdag_classic(dag) +
  theme_dag() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )
```

The first one is A, influencing Z and Y.
The second one is Z, influencing X and Y.

By controlling for Z, we close the path between Z and X **and** the path between A and X!
We will rely on a simple linear regression model:
$$
\text{Y}_i \sim \mathcal{N}(\mu_i, \sigma)
$$
$$
\mu_i = \alpha + \beta_X \text{X}_i + \beta_Z \text{Z}_i
$$

Let's now set the priors.
I'll just start out with normally distributed priors centered at 0 and with a standard deviation of 1.
$$
\alpha \sim \mathcal{N}(0, 1)
$$
$$
\beta_X \sim \mathcal{N}(0, 1)
$$
$$
\beta_Z \sim \mathcal{N}(0, 1)
$$
$$
\sigma \sim \text{Exponential}(1)
$$

To check if these are sensible, we sample from the prior for X and run a prior predictive simulation.
We could do this for all priors, but I think we're okay here if we just do a spot check.

```{r}
#| label: prior_pred_sim
#| cache: false

n <- 20
a <- rnorm(n, 0, 1)
bX <- rnorm(n, 0, 1)

plot(NULL, xlim = c(-10, 10), ylim = c(-10, 10), xlab = "X", ylab = "Y")
Xseq <- seq(from = -10, to = 10, len = 20)
for (i in 1:n) {
  mu <- a[i] + bX[i] * Xseq
  lines(Xseq, mu, lwd = 2, col = 2)
}
```

Okay, since we expect a slightly positive association (we set 0.5 for `b_XY`), and this prior does not rule out such association, it's looking good!

Now let's fit the model using quadratic approximation, `quap`.

```{r}
#| label: quap_1
#| fig-height: 2.5
#| cache: false

m_YXZ <- quap(
  alist(
    Y ~ dnorm(mu, sigma),
    mu <- a + bX * X + bZ * Z,
    a ~ dnorm(0, 1),
    bX ~ dnorm(0, 1),
    bZ ~ dnorm(0, 1),
    sigma ~ dexp(1)
  ),
  data = df
)

precis(m_YXZ)
plot(precis(m_YXZ))
```

## Causal effect of X on Y: Intervention
Let's look at the DAG again.

```{r}
#| label: DAG3
#| echo: false
#| fig-height: 1.5
#| fig-width: 3
#| fig-align: center

dag <- dagitty(
  "dag {
    X -> Y
    Z -> X
    Z -> Y
    A -> Y
    A -> Z
}"
)

coordinates(dag) <- list(
  x = c(X = 1, Y = 3, Z = 2, A = 4),
  y = c(X = 1, Y = 1, Z = 0, A = 0)
)

ggdag_classic(dag) +
  theme_dag() +
  theme(
    plot.background = element_rect(fill = "transparent", color = NA)
  )
```

To estimate the causal effect of X on Y, we want to simulate an intervention on X and see what that does to Y.
We call this "doing" X: 
$$
p(Y|\text{do}(X))
$$

We'll use the posterior distributions from before to do this intervention.
First, we extract the samples from the model we fitted above.
In the next step, we sample 1000 values for Z from our data. 
Next, we simulate Y for X = 0 and X = 1 seprately. 
`with` makes R look into the `post`object that contains the extracted samples so that we don't have to use the `$` operator as often.\
Then, we just compute the contrast between the results by subtracting Y with X = 0 from Y with X = 1.\
Here is the [Link](https://youtu.be/lGR7D45Ww38?si=jpPDjz-q3H-zentB&t=3121) to the part of McElreath's lecture where he explains this.

```{r}
#| label: intervention_X
#| cache: false
#| fig-height: 3.5

post <- extract.samples(m_YXZ)

# sample Z and A from data
n <- 1000
Zs <- sample(df$Z, size = n, replace = T)

# simulate Y for X = 0
YX_0 <- with(post, rnorm(n, a + bX * 0 + bZ * Zs, sigma))

# simulate Y for X = 1
YX_1 <- with(post, rnorm(n, a + bX * 1 + bZ * Zs, sigma))

# contrast
X_1_0 <- YX_1 - YX_0

dens(
  X_1_0,
  lwd = 2,
  col = "lightcoral",
  xlab = "effect of increase in X by 1 on Y"
)
```

The causal effect of X on Y is `r round(mean(X_1_0),2)` in mean (95% HDI: `r round(hdi(X_1_0, credMass = 0.95)[1], 2)`; `r round(hdi(X_1_0, credMass = 0.95)[2], 2)`).
That's close enough to 0.5, which is the value we assigned this relationship when we simulated the data at the beginning of the exercise.

Coming back to the question about the other estimated coefficient, which in this case is just Z: This is not a causal estimate for the association of Z and Y, because this relationship is itself confounded by A.
If we were to interpret this estimate in any causal way here, we'd be in table 2 fallacy land as described in [Westreich & Greenland 2013](https://doi.org/10.1093/aje/kws412)!
